GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2025-07-02 writeUps/paper.tex
finished?? editing

\documentclass[]{article}
\usepackage[letterpaper]{geometry}

\RequirePackage{amsmath,amsfonts,amssymb,amsthm}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
%\usepackage[notablist,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
\usepackage{threeparttable}
\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled

\usepackage[longnamesfirst,sort]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% To set the list of references in 10 point font using natbib.sty

@@

@@

%\begin{aug}
 \author{
 Adam C. Sales\thanks{CONTACT Adam C. Sales. Email: asales@wpi.edu}, Kirk P. Vanacore, and Erin R. Ottmar%\\
 %\affil{Worcester Polytechnic Institute, Worcester, MA, USA}
}

@@

\begin{table}
    \caption{\label{tab:factor} Manipulated factors in simulation study}
  \centering
  %\rowcolors{1}{}{lightgray}
\begin{tabular}{r|ll}%{*{3}{c}}
  \hline
  Factor &\multicolumn{2}{l}{Levels}\\
  \hline
Sample size per condition $n$ &\multicolumn{2}{l}{$100^*,200^*,300^*,400^*,500,600^*,700^*,800^*,900^*,1000$}\\
$\alpha$ &\multicolumn{2}{l}{$0,0.2,0.3^\dagger,0.4^\dagger,0.5,0.6^\dagger,0.7^\dagger,0.8^\dagger,0.9^\dagger,1^\dagger$}\\
The distribution of $\epsilon_i$&\multicolumn{2}{l}{ Normal, Uniform}\\
%$\beta_1$ & \multicolumn{2}{l}{0, 0.3}\\
\multirow{4}{*}{Interactions in \eqref{eq:y-sim}}& None:  $\gamma_1=1/\sqrt{6};\gamma_2=\gamma_3=0$\\ 
& $\bm{x}:\st$: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=1/(2\sqrt{6})$; $\gamma_3=0$\\
& $\bm{x}:Z$: $\gamma_1=1/\sqrt{6};\gamma_2=0$; $\gamma_3=1/(2\sqrt{6})$\\
& $\bm{x}:\st$ \& $\bm{x}:Z$: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=\gamma_3=1/(2\sqrt{6})$\\
% between $\st$ and covariates & No, Yes\\
% Interactions between $Z$ and covariates& No, Yes\\
\hline
\multicolumn{3}{p{4.5in}}{\footnotesize Factors fully crossed except where indicated.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $*$ Only tested with $\alpha=0.5$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $\dagger$ Only tested with $n=500$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}

@@

@@

% \begin{figure}
%   \centering
%   \includegraphics{../simFigs/biasSEbyB1.jpg}
%   \caption{Empirical bias and standard error of M- and Mixture estimates of $\muc1$ as $\alpha$ varies from 0.1 to 1. There were 500 replications for each $n$. Other factors were held fixed at the noted values.}
%   \label{fig:alpha}
% \end{figure}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=5.5in,clip]{../simFigs/boxplotsNew.pdf}
  \caption{Violin and scatter plots of 500 simulation estimation errors for estimators described in Section~\ref{sec:simMods} under varying conditions. For all plots, $n=500$; there were 500 replications for each set of conditions. Annotations indicate the number of extreme outliers excluded from the plotting area.}
  \label{fig:boxplots}
\end{figure}

@@

\subsection{Accounting for Imperfect Compliance: Re-Analysis of the Obstetrics and Periodontal Therapy Study}
\sloppy
Periodontal disease, a weakening of bone and connective tissue supporting teeth, is highly prevalent among American adults and is associated with adverse pregnancy outcomes, including preterm birth \citep{periodontalEpi}.
The Obstetrics and Periodontal Therapy Study \citep[OPTS; ][]{michalowicz2006treatment} investigated the efficacy of nonsurgical periodontal treatment to reduce the preterm birth rate; replication data are available at \url{https://causeweb.org/tshs/obstetrics-and-periodontal-therapy/}.
Pregnant patients at four different hospitals who were at least 16 years old, at less than 16 weeks and 6 days of gestation, and diagnosed with periodontal disease were individually randomized to either receive periodontal treatment during the first 21 weeks of pregnancy ($Z=1$, $n_T=413$) or after delivery ($Z=0$, $n_C=410$).
The main component of the intervention consisted of up to four visits, followed by monthly tooth polishing and other dental care as needed. Not all participants randomized to the treatment condition completed the entire treatment plan; however, 96\% of them received at least some periodontal care during pregnancy.
Patients randomized to the control condition did not have access to the intervention.

@@

\appendix

\section*{Appendix: Proofs and Calculations}
\subsection*{Proof for Lemma \ref{lemma:expectation}}
As a preliminary, note that
% \begin{align*}
%   \EE[\st|\hpp]&=\EE\left\{\EE[\st|\pp,\hpp]|\hpp\right\}\\
%              &=\EE\left\{\EE[\st|\pp]|\hpp\right\}\\
%              &=\EE[\pp|\hpp]=\hpp
% \end{align*}

%Then, note
\begin{align*}
  \EE[Y_C|\pp]&=\EE\left\{\EE[Y_C|\pp,\st]|\pp\right\}\\
             &=\EE\left\{\EE[Y_C|\st]|\pp\right\}\tag*{by \eqref{eq:assumption}}\\
             &=\EE[\muc1\st+\muc0(1-\st)|\pp]\\
             &=\muc1\pp+\muc0(1-\pp)
\end{align*}

Then we have
\begin{equation*}
  \begin{split}
    \EE[Y_C]=&\EE\EE[Y_C|\pp]=\muc1\EE\pp+\muc0(1-\EE\pp)\\
    =&\muc0+\EE\pp(\muc1-\muc0)
    \end{split}
\end{equation*}

Next we have

\begin{align*}
  \EE[Y_C\pp]&=\EE\left\{\EE[Y_C\pp|\pp]\right\}\\
            &=\EE\left\{\pp\EE[Y_C|\pp]\right\}\\
            &=\EE\left\{\pp\left[\muc1\pp+\muc0(1-\pp)\right]\right\}\\
            &=\EE[\pp]\muc0+\EE[\pp^2](\muc1-\muc0)
\end{align*}

In the treatment group, $\st$ is observed, so
\begin{align*}
    \EE[Y_T]=&\mut0+\EE[\st](\mut1-\mut0)\tag*{and}\\
    \EE[\st Y_T]=&\EE[\st]\mut0+\EE[\st^2](\mut1-\mut0)
\end{align*}

Due to Assumption \ref{ass:rand} (randomization), $\EE[Y|Z=0]=\EE[Y_C]$, $\EE[Y|Z=1]=\EE[Y_T]$, $\EE[Y\pp|Z=0]=\EE[Y_C\pp]$ and $\EE[YS|Z=1]=\EE[Y_T\st]$, completing the proof.

\subsection*{Proof for Proposition \ref{prop:reg1}}

Replacing $\sti$ and $\pp$ in \eqref{eq:estEq0} with $\ri$, as in \eqref{eq:ri}, and replacing $\tilde{\Psi}_i$ with $\Psi_i=\begin{psmallmatrix} 1 & 0&1&0\\ 0&1&0&1\\0&0&1&0\\0&0&0&1\end{psmallmatrix}\tilde{\Psi}_i$ gives an equivalent set of estimating equations $\sum_{i=1}^{n_C}\Psi_i=\bm{0}$ with $\Psi_i=$
\begin{equation}\label{eq:estEq1}
\begin{pmatrix}
    Y_i-\muc0-\ri(\muc1-\muc0)-Z_i(\mut0-\muc0)-Z_i\ri(\mut1-\mut0-\muc1+\muc0)\\
    \ri Y_i-\ri\muc0-\ri^2(\muc1-\muc0)-Z_i\ri(\mut0-\muc0)-Z_i\ri^2(\mut0-\mut1-\muc0+\muc1)\\
    Z_iY_i-Z_i\mut0-Z_i\ri (\mut1-\mut0)\\
    Z_i\ri Y_i -Z_i\ri\mut0-Z_i\ri^2(\mut1-\mut0)

\end{pmatrix}
\end{equation}
These are equivalent to the estimating equations for OLS model \eqref{eq:regression0} with $\beta_0=\muc0$, $\beta_1=\muc1-\muc0$, $\beta_2=\mut0-\muc0$, and $\beta_3=\mut1-\mut0-\muc1-\muc0$.
Therefore, under standard OLS regularity conditions the estimated parameter vector $\bm{\hat{\beta}}$ is consistent, completing the proof.

\subsection*{A Stronger Version of Proposition \ref{prop:reg2} and a Proof}

\begin{prop}\label{prop:interactions}
  Say, for $i=1,\dots,n$, principal scores $\ppi$ are generated as \eqref{eq:pscore}, with parameters $\bm{\alpha}$ identified and consistently estimable with M-estimation, and there exist $\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$, $\bm{\gamma_1}$, $\bm{\gamma_2}$, $\bm{\gamma_3}$ and $\bm{\gamma_4}$ such that $\{Y_i,Z_i,\sti,\bxy_i\}_{i=1}^n$ are independent and identically distributed with
  \begin{equation}\label{eq:interaction}
    \begin{split}
    \EE[Y_i|\st,Z,\bx]=&\beta_0+\beta_1\sti+\beta_2 Z_i+\beta_3Z_i\sti\\
    &+\bm{\gamma_1}'\bxy_i+\bm{\gamma_2}'\bxy_i\sti+
    \bm{\gamma_3}'\bxy Z_i+\bm{\gamma_4}'\bxy_i Z_i\sti
    \end{split}
  \end{equation}

  Then, under Assumptions \ref{ass:sm}, \ref{ass:rand}, and \ref{ass:vps}, if $\ppi$ is linearly independent of $\bxy$, a researcher may follow the following procedure to estimate principal effects:
  \begin{enumerate}
  \item Estimate principal scores by fitting model \eqref{eq:pscore} to data from the treatment group
  \item Replace $\sti$ with $\ri$ (as defined in \ref{eq:ri}) in model \eqref{eq:interaction} and fit with OLS
  \item Estimate principal effects as:
   \begin{equation}\label{eq:prinEffEstApp}
  \begin{split}
    \heff0_{int}&\equiv \hat{\beta}_2+\bm{\hat{\gamma}_3}'\overline{\bxy}_{Z=1,S=0}\\
    \heff1_{int}&\equiv \hat{\beta}_2+\hat{\beta}_3+(\bm{\hat{\gamma}_3}+\bm{\hat{\gamma}_4})'\overline{\bxy}_{Z=1,S=1}
  \end{split}
   \end{equation}
   where $\overline{\bxy}_{Z=1,S=0}$ and $\overline{\bxy}_{Z=1,S=0}$ are the vector of covariate sample means for the subsets of subjects with $Z=1$ and $S=0$ or $S=1$, respectively.
  \end{enumerate}
  Then $\heff0_{G}$ and $\heff1_{int}$ are M-estimators. If the estimating equations for \eqref{eq:prinEffEstApp} are each bounded by an integrable function of $\{\bm{Y},\bxy, \bm{S},\pp,\bm{Z}\}$ that does not depend on $\{\bm{\beta},\bm{\gamma}\}$, then $\heff0_{int}\rightarrow_p\eff0$ and $\heff1_{int}\rightarrow_p\eff1$ as $n\rightarrow\infty$.

  If the parameter estimates of the principal score model are asymptotically normal, second partial derivatives of the estimating equations for \eqref{eq:prinEffEstApp} are bounded by an integrable function of the data for values of $\{\bm{\beta},\bm{\gamma}\}$ in a neighborhood of their probability limits, and the sandwich components of \eqref{eq:sandwich}, $A$ and $B$, exist and are finite, and if $B$ is non-singular, then $\heff0_{int}$ and $\heff1_{int}$ are jointly asymptotically normal, with a variance of the form \eqref{eq:sandwich}.
\end{prop}

Equation \eqref{eq:interaction} implies Assumption \ref{ass:rci} with $\bm{\gamma_2}=\bm{\gamma_3}=\bm{\gamma_4}=0$.

\begin{proof}
First of all, by \eqref{eq:interaction},
\begin{equation*}
\begin{split}
  \EE[Y_T-Y_C|\st=0]&\\
  =&\EE[Y|Z=1,\st=0]-\EE[Y|Z=1,\st=0]\\
  =&\beta_2+\bm{\gamma_3}'\EE[\bxy|\st=0]
\end{split}
\end{equation*}
and
\begin{equation*}
\begin{split}
  \EE[Y_T-Y_C|\st=1]&\\
  =&\EE[Y|Z=1,\st=1]-\EE[Y|Z=1,\st=1]\\
  =&\beta_2+\beta_3+(\bm{\gamma_3}'+\bm{\gamma_4}')\EE[\bxy|\st=1]
\end{split}
\end{equation*}
Furthermore, $\overline{\bxy}_{Z=1,S=0}\rightarrow \EE[\bxy|\st=0]$ and $\overline{\bxy}_{Z=1,S=1}\rightarrow \EE[\bxy|\st=1]$ as $n\rightarrow \infty$.

We will show that the estimated coefficients from model \eqref{eq:interaction}, but with $R$ replacing $\st$, fit with OLS, are consistent for $\bm{\beta}$ and $\bm{\gamma}$ from \eqref{eq:interaction}.

First, note that $\EE[S]=\EE\EE[S|\bx]=\EE[\pp]$ and
\begin{align*}
  \EE[\bx S]&=\EE[\bx S|Z=1]=\EE[\bx S|Z=0] \mbox{ (due to randomization)}\\
  &=\EE[\bx\EE[S|\bx]|Z=0]=\EE[\bx\pp|Z=0]=\EE[\bx\pp]
\end{align*}
implying that, according to \eqref{eq:interaction},
\begin{equation*}
  \begin{split}
    \EE[Y|\bx,Z=0]&=\beta_0+\beta_1\pp+\bm{\gamma_1}'\bxy+\bm{\gamma_2}'\bxy\pp\\
                  &=\beta_0+\beta_1R+\bm{\gamma_1}'\bxy+\bm{\gamma_2}'\bxy R\\
    \EE[Y|\bx,S,Z=1]&=\beta_0+\beta_2+(\beta_1+\beta_3)S+(\bm{\gamma_1}+\bm{\gamma_3})\bxy+(\bm{\gamma_2}+\bm{\gamma_4})\bxy S\\
    &=\beta_0+\beta_2+(\beta_1+\beta_3)R+(\bm{\gamma_1}+\bm{\gamma_3})\bxy+(\bm{\gamma_2}+\bm{\gamma_4})\bxy R
  \end{split}
\end{equation*}
Therefore,
\begin{align*}
  \EE[Y]=&\EE[Y|Z=0]+\EE[Z]\left\{\EE[Y|Z=1]-\EE[Y|Z=0]\right\}\\
  =&\beta_0+\beta_1\EE[R]+\beta_2\EE[Z]+\beta_3\EE[ZR]+\bm{\gamma_1}'\EE[\bxy]+\bm{\gamma_2}'\EE[\bxy R]\\
  &+\bm{\gamma_3}'\EE[\bxy Z]+\bm{\gamma_4}\EE[\bxy RZ]
\end{align*}

Analogous reasoning leads to expressions for $\EE[RY]$, $\EE[\bxy Y]$, $\EE[ZY]$,  $\EE[ZRY]$, $\EE[\bxy YZ]$, and $\EE[\bxy RZY]$.
These, in turn, give rise to estimating equations
\begin{equation*}
\begin{split}
  &\psi_i=\\ &\begin{pmatrix}
  Y_i\\
  \phantom{}\\
  Y_i\ri\\
    \phantom{}\\
  Y_iZ_i\\
  \phantom{}\\
  Y_iZ_i\ri\\
  \phantom{}\\
  Y_i\bxy_i\\
  \phantom{}\\
  Y_i\bxy_i\ri\\
  \phantom{}\\
  Y_i\bxy_iZ_i\\
  \phantom{}\\
  Y_i\bxy_iZ_i\ri\\
  \phantom{}\end{pmatrix} - \begin{pmatrix*}[l]
    \beta_0+\beta_1R_i+\beta_2Z_i+\beta_3Z_iR_i\\
    \quad+\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\\
    \beta_0\ri+\beta_1R_i^2+\beta_2Z_i\ri+\beta_3Z_iR_i^2\\
    \quad+\ri\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\\
    (\beta_0+\beta_2)Z_i+(\beta_1+\beta_3)R_iZ_i\\
    \quad+Z_i\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\\
    \beta_0Z_i\ri+(\beta_1+\beta_3)Z_iR_i^2+\beta_2Z_i\ri\\
    \quad+Z_iR_i\left\{\bm{\gamma_1}'+\bm{\gamma_3}'+R_i(\bm{\gamma_2}'+\bm{\gamma_4}')\right\}\bxy \\
    \beta_0\bxyt+\beta_1R_i\bxyt+\beta_2Z_i\bxyt+\beta_3Z_iR_i\bxyt\\
    \quad {}+\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\bxyt\\
    \beta_0\ri\bxyt+\beta_1R_i^2\bxyt+\beta_2Z_i\ri\bxyt+\beta_3Z_iR_i^2\bxyt\\
    \quad {}+\ri\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\bxyt\\
        (\beta_0+\beta_2)Z_i\bxyt+(\beta_1+\beta_3)R_iZ_i\bxyt\\
    \quad {}+Z_i\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\bxyt\\
\beta_0\ri Z_i\bxyt+\beta_1R_i^2Z_i\bxyt+\beta_2\ri Z_i\bxyt+\beta_3Z_iR_i^2\bxyt\\
    \quad {}+\ri Z_i\left\{\bm{\gamma_1}'+R_i\bm{\gamma_2}'+Z_i\bm{\gamma_3}'+R_iZ_i\bm{\gamma_4}'\right\}\bxy\bxyt\\
  \end{pmatrix*}
  \end{split}
\end{equation*}
with $\EE[\psi_i]=0$.
These are the estimating equations for the regression model \eqref{eq:interaction}, with $R$ replacing $\st$, fit by OLS.
Consistency and asymptotic normality follow from theorems 7.8.1 and 7.8.2, respectively, of \citet{boosStefanskiBook}
\end{proof}

\subsection*{Sandwich Matrix Calculations}

Here we will derive the sandwich variance-covariance matrix for the \textsc{geepers} estimate without interactions between $\bx$ and either $Z$ or $\st$---i.e., with $\bm{\gamma_2}=\bm{\gamma_3}=\bm{\gamma_4}=0$ in the notation of \eqref{eq:interaction}---and estimating principal scores using a generalized linear model.

We propose estimating principal effects in two stages.
First, fit the model
\begin{equation}\label{eq:psMod}
  \ppi=Pr(\sti=1|\bxsi)=f(\bm{\alpha}'\bxsit)
\end{equation}
for some inverse link function $f(\cdot)$, where $\bxsit=[1,\bxsi]$, using (observed) values from the treatment group, and estimating $\hat{\alpha}$.
Then let
\begin{equation}\label{eq:ps}
  \hat{p}_i=f(\bm{\hat{\alpha}}'\bxsit)
\end{equation}
for all subjects in the experiment.

Finally, fit model
\begin{equation}\label{eq:regression}
  Y_i=\beta_0+\beta_1r_i+\beta_2Z_i+\beta_3Z_i\ri+\bm{\gamma}'\bxy_i+\epsilon_i
\end{equation}
to estimate $\bm{\beta}$ and hence principal effects, where $\bm{x}_i$ is a set of covariates predictive of $Y$ within principal strata.
Let $\bm{\beta}=[\beta_0,\beta_1,\beta_2,\beta_3,\bm{\gamma}']'$.

\sloppy
Following \eqref{eq:stacked}, let $\blam(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\alpha},\bm{\beta})=\begin{pmatrix} Z_i\Omega(\bxsi,S_i;\bm{\alpha})\\ \Psi(\bxy_i,\bxsi,Y_i,Z_i,\ri(\bm{\alpha});\bm{\beta})\end{pmatrix}$, the stacked estimating equations of \eqref{eq:psMod} and \eqref{eq:regression}.
Going forward, for the sake of brevity, we will write $\blam_i(\bm{\alpha},\bm{\beta})=\blam(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\hat{\alpha}},\bm{\hat{\beta}})$, where dependence on the data for $i$ is captured in the subscript $i$, with similar meanings for $\Psi_i(\bm{\alpha},\bm{\beta})$ and $\omega_i(\bm{\alpha})$.
%Then let $\blamh_i=\blam_i(Z_i,S_i,\bxsi,\bxy_i,Y_i;\bm{\hat{\alpha}},\bm{\hat{\beta}})$ be the estimating equations evaluated at the estimated parameters, and let

%\begin{equation*}
%\dblam_i=\frac{\partial}{\partial [\alpha,\beta]'} \blam_i\Bigr|_{\substack{\bm{\alpha}=\bm{\hat{\alpha}}\\\bm{\beta}=\bm{\hat{\beta}}}}
%\end{equation*}
%the derivative matrix of $\blam$ evaluated at the estimated parameters.

The variance-covariance matrix for $\bm{\hat{\alpha}}$ and $\bm{\hat{\beta}}$ can be estimated as:
\begin{equation*}
  \widehat{var}\left([\bm{\hat{\alpha}}',\bm{\hat{\beta}}']'\right)=A^{-1}BA^{-t}
\end{equation*}
where
\begin{equation*}%\label{eq:Amat}
  A=\sum_i \frac{\partial}{\partial [\bm{\alpha},\bm{\beta}]'} \blam_i\Bigr|_{\substack{\bm{\alpha}=\bm{\hat{\alpha}}\\\bm{\beta}=\bm{\hat{\beta}}}}
\end{equation*}
and
\begin{equation*}%\label{eq:Bmat}
  B=\sum_i \blam_i(\bm{\hat{\alpha}},\bm{\hat{\beta}})\blam_i(\bm{\hat{\alpha}},\bm{\hat{\beta}})'
\end{equation*}

Following \citet[][p. 373]{carroll2006measurement}, we %separate the estimating equations $\Psi$ into $\phi$ and $\psi$, estimating equations for models \eqref{
can decompose the matrices into diagonal elements
\begin{equation*}
    \begin{split}
        A_{1,1}&=\sum_i \partial \Omega_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
        A_{2,2}&=\sum_i\partial\Psi_i/\partial \bm{\beta}|_{\bm{\beta}=\bm{\hat{\beta}}}\\
        B_{1,1}&=\sum_i\Omega_i(\bm{\hat\alpha})\Omega_i(\bm{\hat\alpha})'\\
        B_{2,2}&=\sum_i \Psi_i(\bm{\hat\alpha},\bm{\hat\beta})\Psi_i(\bm{\hat\alpha},\bm{\hat\beta})'
    \end{split}
\end{equation*}
 that pertain to the parameter sets $\bm{\alpha}$ and $\bm{\beta}$ and the estimating equations for models \eqref{eq:psMod} and  \eqref{eq:regression}, respectively, and
 \begin{equation*}
     \begin{split}
         A_{21}&=\sum_i\partial\Psi_i/\partial \bm{\alpha}|_{\bm{\alpha}=\bm{\hat{\alpha}}}\\
         B_{12}=B_{21}'&=\sum_i \Omega_i(\bm{\hat{\alpha}})\Psi_i(\bm{\hat{\alpha}},\bm{\hat{\beta}})'
     \end{split}
 \end{equation*}
 which capture the dependence of model \eqref{eq:regression} on the parameters $\bm{\alpha}$ from \eqref{eq:psMod} and the covariance between the estimating equations of the two models.

 The sub-matrix $A_{12}=\sum_i \partial \Omega_i/\partial \bm{\beta}=0$, since \eqref{eq:psMod} does not depend on $\bm{\beta}$.

The diagonal matrices $A_{1,1}$ and $A_{2,2}$ and $B_{1,1}$ and $B_{2,2}$ are all the typical ``bread'' and ``meat'' matrices from M-estimation of generalized linear models and OLS.
%In practice, we use the estimates from the \texttt{sandwich} package in \texttt{R}, adjusted in two ways: first, the function \texttt{bread} actually gives $A^{-1}$, not $A$; second, we must pay careful attention to sample sizes, since the sample size for \eqref{eq:psMod} includes only treated observations (with observed $S$) and \eqref{eq:regression} contains all observations.
Calculation of the matrices $B_{12}$ and $B_{21}$ is straightforward after vectors $\Omega_i(\bm{\hat{\alpha}})$ and $\Psi_i(\bm{\hat{\alpha}},\bm{\hat{\beta}})$ have been calculated.
Some specialized calculation is necessary for matrix $A_{21}$.

\subsection*{$A_{21}$ Matrix}
The estimating equations for the regression \eqref{eq:regression} are
\begin{equation}\label{eq:eeOLS}
  \psi(Y_i,\bxy_i,\bxsi,\bm{\beta},\alpha)=X_iY_i-X_iX_i'\bm{\beta}
\end{equation}
Where $X_i=[1,r_i,Z_i,r_iZ_i,\bxyt_i]'$.
In other words,
\begin{align*}
  \psi(Y_i,&\bxy_i,\bxsi,\bm{\beta},\bm{\alpha})=\\
  &\left\{
  \begin{array}{l}
    Y_i-\left(\beta_0+\beta_1r_i+\beta_2Z_i+\beta_3Z_ir_i+\bm{\gamma}'\bxy_i\right)\\
    r_iY_i-r_i\left(\beta_0+\beta_2Z_i+\bm{\gamma}'\bxy_i\right)-r_i^2\left(\beta_1+\beta_3Z_i\right)\\
    Z_iY_i-Z_i\left(\beta_0+\beta_1r_i+\beta_2Z_i+\beta_3Z_ir_i+\bm{\gamma}'\bxy_i\right)\\
    Z_ir_iY_i-Z_ir_i\left(\beta_0+\beta_2Z_i+\bm{\gamma}'\bxy_i\right)-Z_ir_i^2\left(\beta_1+\beta_3Z_i\right)\\
    \bxy_iY_i-\bxy_i\left(\beta_0+\beta_1r_i+\beta_2Z_i+\beta_3Z_ir_i+\bm{\gamma}'\bxy_i\right)
  \end{array}
  \right\}
\end{align*}
(noting that $Z^2=Z$).
These depend on $\bm{\alpha}$ when $r_i=p_i$, i.e., when $Z_i=0$.% or $S_i$ is missing.
Then note that, following \eqref{eq:ps}, and letting $\eta_i=\bm{\alpha}'\bxsit$
\begin{equation}\label{eq:derivP}
  \frac{\partial p_i}{\partial \bm{\alpha}'}=f'(\eta_i)\frac{\partial \eta_i}{\partial \bm{\alpha}'}=f'(\eta_i)\bxsitp
\end{equation}
and that
\begin{equation}
  \frac{\partial p_i^2}{\partial \bm{\alpha}'}=2p_i\frac{\partial p_i}{\partial \bm{\alpha}'}=2f(\eta)f'(\eta_i)\bxsitp=2p_if'(\eta_i)\bxsitp
\end{equation}

>>>>

\documentclass[]{article}
\usepackage[letterpaper]{geometry}
\usepackage{xr}
\externaldocument{writeUps/appendix}
\RequirePackage{amsmath,amsfonts,amssymb,amsthm}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage[caption=false]{subfig}
\usepackage[doublespacing]{setspace}

\usepackage[longnamesfirst,sort]{natbib}
\bibpunct[, ]{(}{)}{;}{a}{,}{,}%
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% To set the list of references in 10 point font using natbib.sty

@@

\usepackage{endfloat}% To `separate' figures and tables from text if required

@@

%\begin{aug}
%  \author{
%  Adam C. Sales\thanks{CONTACT Adam C. Sales. Email: asales@wpi.edu}, Kirk P. Vanacore, and Erin R. Ottmar%\\
%  %\affil{Worcester Polytechnic Institute, Worcester, MA, USA}
% }

@@

\begin{table}
    \caption{\label{tab:factor} Manipulated factors in simulation study}
  \centering
  %\rowcolors{1}{}{lightgray}
\begin{tabular}{r|ll}%{*{3}{c}}
  \hline
  Factor &\multicolumn{2}{l}{Levels}\\
  \hline
Sample size per condition $n$ &\multicolumn{2}{l}{$100^*,200^*,300^*,400^*,500,600^*,700^*,800^*,900^*,1000$}\\
$\alpha$ &\multicolumn{2}{l}{$\rd{0},0.2,0.3^\dagger,0.4^\dagger,0.5,0.6^\dagger,0.7^\dagger,0.8^\dagger,0.9^\dagger,1^\dagger$}\\
The distribution of $\epsilon_i$&\multicolumn{2}{l}{ Normal, 
\rd{Uniform}}\\
%$\beta_1$ & \multicolumn{2}{l}{0, 0.3}\\
\multirow{4}{*}{Interactions in \eqref{eq:y-sim}}& None:  $\gamma_1=1/\sqrt{6};\gamma_2=\gamma_3=0$\\ 
& \rd{$\bm{x}:\st$}: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=1/(2\sqrt{6})$; $\gamma_3=0$\\
& \rd{$\bm{x}:Z$}: $\gamma_1=1/\sqrt{6};\gamma_2=0$; $\gamma_3=1/(2\sqrt{6})$\\
& \rd{$\bm{x}:\st$ \& $\bm{x}:Z$}: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=\gamma_3=1/(2\sqrt{6})$\\
% between $\st$ and covariates & No, Yes\\
% Interactions between $Z$ and covariates& No, Yes\\
\hline
\multicolumn{3}{p{4.5in}}{\footnotesize Factor levels in red violate assumptions of \geepers, \pmm, or both. Factors fully crossed except where indicated.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $*$ Only tested with $\alpha=0.5$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $\dagger$ Only tested with $n=500$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}

@@

For reference, in the application of Section~\ref{sec:opts}, there were two covariates. In the model for principal scores, their standardized coefficients were roughly 0.46 and 0.62, and the AUC was 0.68. In the application of Section~\ref{sec:fh2t}, the principal score models (which were far more complex than either the simulation model or the model in Section \ref{sec:opts}) had AUCs between 0.745 and 0.795.

@@

\begin{figure}
  \centering
  \includegraphics{../simFigs/biasSEbyB1.jpg}
  \caption{Empirical bias and standard error of M- and Mixture estimates of $\muc1$ as $\alpha$ varies from 0.1 to 1. There were 500 replications for each $n$. Other factors were held fixed at the noted values.}
  \label{fig:alpha}
\end{figure}

@@

% \begin{figure}[!ht]
%   \centering
%   \includegraphics[width=5.5in,clip]{../simFigs/boxplotsNew.pdf}
%   \caption{Violin and scatter plots of 500 simulation estimation errors for estimators described in Section~\ref{sec:simMods} under varying conditions. For all plots, $n=500$; there were 500 replications for each set of conditions. Annotations indicate the number of extreme outliers excluded from the plotting area.}
%   \label{fig:boxplots}
% \end{figure}

@@

\subsection{Accounting for Imperfect Compliance: Re-Analysis of the Obstetrics and Periodontal Therapy Study}\label{sec:opts}
\sloppy
Periodontal disease, a weakening of bone and connective tissue supporting teeth, is highly prevalent among American adults and is associated with adverse pregnancy outcomes, including preterm birth \citep{periodontalEpi}.
The Obstetrics and Periodontal Therapy Study \citep[OPTS; ][]{michalowicz2006treatment} investigated the efficacy of nonsurgical periodontal treatment to reduce the preterm birth rate; replication data are available at \url{https://causeweb.org/tshs/obstetrics-and-periodontal-therapy/}.
Pregnant patients at four different hospitals who were at least 16 years old, at less than 16 weeks and 6 days of gestation, and diagnosed with periodontal disease were individually randomized to either receive periodontal treatment during the first 21 weeks of pregnancy ($Z=1$, $n_T=413$) or after delivery ($Z=0$, $n_C=410$).
The main component of the intervention consisted of up to four visits, followed by monthly tooth polishing and other dental care as needed. Not all participants randomized to the treatment condition completed the entire treatment plan; however, 96\% of them received at least some periodontal care during pregnancy.
Patients randomized to the control condition did not have access to the intervention.

@@
